Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/overview.tex
overview section is now sec:overview
removing obsolete text from comments

\section{Overview}
\label{sec:examplea}

@@

\begin{comment}

An astute programmer, having noticed that \texttt{quickselect} can be staged in
this fashion, might try to split it into a pair of functions, one which performs
all the work depending only on \texttt{l} (the first stage), and one which uses
that partial result and \texttt{k} to compute the element with
rank \texttt{k} in \texttt{l}.

Intuitively, \texttt{l} determines the result of all calls to
\texttt{part}ition, and \texttt{k} only determines which calls are made. So we
can preprocess \texttt{l} by recursively dividing it into halves smaller and
greater than the pivot---that is, building a binary search tree. Then, once we
have \texttt{k}, we can recur on this tree, choosing whichever branch has the
\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. And because
\texttt{part}ition contains no second-stage code, we can run it entirely in the
first stage.

We have implemented this splitting of quickselect in \ref{fig:qs-split}.
\texttt{qSelect1} builds a binary search tree from the list \texttt{l}, and
\texttt{qSelect2} takes such a tree and a rank \texttt{k} and computes the
answer. This allows us to efficiently perform many order statistics queries on
\texttt{l} by caching the tree and reusing it for many different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$:
%
\begin{lstlisting}
let b = qSelect1 l in
  qSelect2 b @$k_1$@
  qSelect2 b @$k_2$@
   @$\vdots$@ 
  qSelect2 b @$k_m$@.
\end{lstlisting}

Assuming \texttt{l} contains $n$ elements, this optimization changes the
asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$ to
$\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a significant improvement.

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, given a program $e$ in \lang,
produces an equivalent pair of programs which correspond precisely to the two
stages of computation in $e$.
(Splitting is always possible because the staging annotations in $e$ are
consistent, because $e$ is well-typed in \lang.) In the case of
\texttt{qsStaged}, our splitting algorithm produces the algorithm described
above.
\end{comment}
%\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.

%Because the tree passes information across the stage boundary, we call it the
%\emph{boundary data structure}.

%Note that the desired optimized code shows above is intellectually more
%sophisticated than the code that we have started with: the optimized code is
%able to create a data structure, a balanced binary tree augmented with indexing
%information, and use a binary search technique over this tree to compute the
%result asymptotically more efficiently.

%In fact, based our teaching experience, we can imagine this kind of problem to
%be a moderately difficult exam question in an undergraduate algorithms class, as
%it not only requires understanding of data structures such as binary search
%trees but also requires modifying them to augment with indexing information to
%support rank-based search.

%(I also want to make it clear that recognizing \lang's appropriateness for this
%is itself is a contribution.)

>>>>

\section{Overview}
\label{sec:overview}

@@
